---
title: "Final report"
date: "2024-12-17"
output: html_document
---

```{r library}
library(tidyverse)
library(janitor)
library(ggplot2)
library(GGally)
library(dplyr)
library(gridExtra)
library(survival)
library(car)
library(PerformanceAnalytics)
library(glmnet)
library(survival)
library(tidyr)
set.seed(1)
```

```{r data input}
## 1
## load the data
project_2_data = 
  read_csv("Project_2_data.csv", na = c("NA", "", ".")) |>
  janitor::clean_names()  |>
  mutate(status = ifelse(status == "Dead", 1, 0))
```

```{r Descriptive_data}
## Summary statistics for numeric variables
numerical_vars <- project_2_data[c("age","tumor_size","regional_node_examined","reginol_node_positive","survival_months")]
summary(numerical_vars)

## Summary for categorical variables
project_2_data %>%
  dplyr::select(where(~!is.numeric(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "category") %>%
  count(variable, category) %>%
  group_by(variable) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  print()

```
```{r preprocesssing}
## 2
project_2_data <-
  na.omit(project_2_data)|>
  mutate(
    t_stage = case_when(
    t_stage == "T1" ~ 1,
    t_stage == "T2" ~ 2,
    t_stage == "T3" ~ 3,
    t_stage == "T4" ~ 4,
    TRUE ~ NA_real_),
    n_stage = case_when(
    n_stage == "N1" ~ 1,
    n_stage == "N2" ~ 2,
    n_stage == "N3" ~ 3,
    TRUE ~ NA_real_),
    differentiate = case_when(
    differentiate == "Well differentiated" ~ 1,
    differentiate == "Moderately differentiated" ~ 2,
    differentiate == "Poorly differentiated" ~ 3,
    differentiate == "Undifferentiated" ~ 4,
    TRUE ~ NA_real_),
    grade = case_when(
    grade == "anaplastic; Grade IV" ~ 4,
    grade == "3" ~ 3,
    grade == "2" ~ 2,
    grade == "1" ~ 1,
    TRUE ~ NA_real_),
    a_stage_regional = ifelse(a_stage == "Regional", 1, 0),
    estrogen_status = ifelse(estrogen_status == "Positive", 1, 0),
    progesterone_status = ifelse(progesterone_status == "Positive", 1, 0)
    ) |>
  dplyr::select(-a_stage)

#QQ plots
ggplot(project_2_data, aes(sample = survival_months)) +
  stat_qq() +
  stat_qq_line()

ggplot(project_2_data, aes(sample = tumor_size)) +
  stat_qq() +
  stat_qq_line()

ggplot(project_2_data, aes(sample = age)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot for age", x = "Theoretical Quantiles", y = "Sample Quantiles")

ggplot(project_2_data, aes(sample = regional_node_examined)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(title = "QQ Plot for regional_node_examined", x = "Theoretical Quantiles", y = "Sample Quantiles")

ggplot(project_2_data, aes(sample =reginol_node_positive)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(title = "QQ Plot for reginol_node_positive", x = "Theoretical Quantiles", y = "Sample Quantiles")

project_2_data <- project_2_data %>%
  mutate(sqrt_survival_months = sqrt(survival_months))

ggplot(project_2_data, aes(x = sqrt_survival_months)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "black") +
  labs(title = "Square root-Transformed Survival Months", x = "Square Root(Survival Months)", y = "Frequency")

ggplot(project_2_data, aes(sample = sqrt_survival_months)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(
    title = "QQ Plot for Square Root-Transformed Survival Months",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

project_2_data <- project_2_data %>%
  mutate(log_survival_months = log(survival_months + 1), 
         log_tumor_size = log(tumor_size + 1))

# Visualization after log transformation
ggplot(project_2_data, aes(x = log_survival_months)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "black") +
  labs(title = "Log-Transformed Survival Months", x = "Log(Survival Months)", y = "Frequency")

## QQ-plot in survival months after log transformation
ggplot(project_2_data, aes(sample = log_survival_months)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(
    title = "QQ Plot for Log-Transformed Survival Months",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )
### do not involve transformation for survival months

# Square root transformation for Tumor Size
project_2_data <- project_2_data %>%
  mutate(sqrt_tumor_size = sqrt(tumor_size))

# Visualization after square root transformation
ggplot(project_2_data, aes(x = sqrt_tumor_size)) +
  geom_histogram(bins = 30, fill = "purple", color = "black") +
  labs(title = "Square Root Transformed Tumor Size", x = "Sqrt(Tumor Size)", y = "Frequency")

# QQ Plot: Square root-transformed data
ggplot(project_2_data, aes(sample = sqrt_tumor_size)) +
  stat_qq() +
  stat_qq_line(color = "green") +
  labs(
    title = "QQ Plot for Square Root Transformed Tumor Size",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )
## Square root transformation for tumor size is needed

# Square root transformation for regional_node_positive
project_2_data <- project_2_data %>%
  mutate(sqrt_reginol_node_positive = sqrt(reginol_node_positive))

# Visualization after square root transformation
ggplot(project_2_data, aes(x = sqrt_reginol_node_positive)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Square Root Transformed Regional Node Positive", x = "Sqrt(Regional Node Positive)", y = "Frequency")

# QQ Plot: Square root-transformed data
ggplot(project_2_data, aes(sample = sqrt_reginol_node_positive)) +
  stat_qq() +
  stat_qq_line(color = "green") +
  labs(
    title = "QQ Plot for Square Root Transformed Regional Node Positive",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

# Inverse transformation for regional_node_positive
project_2_data <- project_2_data %>%
  mutate(inverse_reginol_node_positive = 1/reginol_node_positive)

# Visualization after square root transformation
ggplot(project_2_data, aes(x = inverse_reginol_node_positive)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Square Root Transformed Regional Node Positive", x = "Sqrt(Regional Node Positive)", y = "Frequency"
       )

# QQ Plot: Square root-transformed data
ggplot(project_2_data, aes(sample = inverse_reginol_node_positive)) +
  stat_qq() +
  stat_qq_line(color = "green") +
  labs(
    title = "QQ Plot for Square Root Transformed Regional Node Positive",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

# Log transformation for regional_node_positive
project_2_data <- project_2_data %>%
  mutate(log_reginol_node_positive = log(reginol_node_positive + 1), # Adding 1 to avoid log(0)
         log_reginol_node_positive = log(reginol_node_positive + 1))

# Visualization after log transformation
ggplot(project_2_data, aes(x = log_reginol_node_positive)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "black") +
  labs(title = "Log-Transformed Regional Node Positive", x = "Log(Survival Months)", y = "Frequency")

# QQ Plot: Square log-transformed data
ggplot(project_2_data, aes(sample = log_reginol_node_positive)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(
    title = "QQ Plot for Log-Transformed Regional Node Positive",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )


#plot after balance dead and alive
ggplot(project_2_data, aes(x = survival_months)) +
  geom_histogram(bins = 30, color = "black")

ggplot(project_2_data, aes(x = status)) +
  geom_bar()


dead <- project_2_data |>
  filter(status == 1)

alive <- project_2_data |>
  filter(status == 0) |>
  sample_n(nrow(dead))

balance_df<-
  bind_rows(dead, alive)

#histogram and bar plot
ggplot(balance_df, aes(x = age)) +
  geom_histogram(bins = 30, color = "black")

ggplot(balance_df, aes(x = race)) +
  geom_bar()

ggplot(balance_df, aes(x = marital_status)) +
  geom_bar()

ggplot(balance_df, aes(x = t_stage)) +
  geom_bar()

ggplot(balance_df, aes(x = n_stage)) +
  geom_bar()

ggplot(balance_df, aes(x = x6th_stage)) +
  geom_bar()

ggplot(balance_df, aes(x = differentiate)) +
  geom_bar()

ggplot(balance_df, aes(x = grade)) +
  geom_bar()

ggplot(balance_df, aes(x = a_stage_regional)) +
  geom_bar()

ggplot(balance_df, aes(x = tumor_size)) +
  geom_histogram(bins = 30, color = "black")

ggplot(balance_df, aes(x = log2(tumor_size+1))) +
  geom_histogram(bins = 30, color = "black") #best

ggplot(balance_df, aes(x = sqrt(tumor_size))) +
  geom_histogram(bins = 30, color = "black")

ggplot(balance_df, aes(x = log2(tumor_size+1))) +
  geom_boxplot()#best1

ggplot(balance_df, aes(x = sqrt(tumor_size))) +
  geom_boxplot()
```


```{r Multicollinearity }
cor_matrix <- cor(balance_df[, sapply(balance_df, is.numeric)])
print(cor_matrix)


cortable<-balance_df|>
  dplyr::select(-race,-x6th_stage, -marital_status)
chart.Correlation(cortable, histogram=TRUE, pch=19)

balance_df <-
  balance_df|>
  dplyr::select(-differentiate)
```
differentiate and grade are 100% correlated
need to decided between t_stage and tumor_size, and  n_stage and reginol_node_exam

```{r scatter plot }
### 3

# scatter plot for Y and numerical variable 
numerical_vars <- c("age","tumor_size","regional_node_examined","reginol_node_positive")
plot_list1 <- list()
for (var in numerical_vars) {
  p<- ggplot(project_2_data, aes_string(x = var, y = "survival_months")) + 
    geom_point(color = "blue", alpha = 0.6) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = paste (var, "vs Survival Months"), x = var, y = "Survival Months")+    theme_minimal() 
  plot_list1[[var]] <- p
}
grid.arrange(grobs = plot_list1, ncol = 2)

### box plot for Y and categorical variable
categorical_vars <- c("race","marital_status","t_stage","n_stage","x6th_stage","differentiate","grade","a_stage","estrogen_status", "progesterone_status")
plot_list2 <- list()
for (var in categorical_vars) {
  m<- ggplot(project_2_data, aes_string(x = var, y = "survival_months")) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    labs(title = paste (var, "vs Survival Months"), x = var, y = "Survival Months") + 
    theme_minimal()
  plot_list2[[var]] <- m
}
grid.arrange(grobs= plot_list2, ncol = 5)

### 1. We get the summary statistics and make the table in excel to have a better format
### 2. We suggest tumor size to have a square root transformation, further transformation are still needed after finish the modeling using box-cox 
### 3. there are potential outliers identified but we might ignore them because they are quite limited and we have large sample size
```

# Logistic model
```{r setup}
library(tidyverse)
library(survival)
library(caret)
#car package for vif
library(car)
#MASS package for var selection
library(MASS)

data <- read.csv("Project_2_data.csv")
view(data)

sampledf <- data |>
  mutate(Status = ifelse(Status == "Dead", 1, 0))

dead <- sampledf |>
  filter(Status == 1)

alive <- sampledf |>
  filter(Status == 0) |>
  sample_n(nrow(dead))

final<-bind_rows(dead, alive) |>
  dplyr::select(-Survival.Months)

logit_model <- glm(Status ~ ., data = final, family = binomial())

summary(logit_model)

predictions <- predict(logit_model, type = "response")

final$predicted_classes <- ifelse(predictions > 0.5, 1, 0)

confusionMatrix(as.factor(final$predicted_classes), as.factor(final$Status))
```
## Assumptions
```{r multicollinearity}
#vif(logit_model)
# there are aliased coefficients in the model > means some vars have perfect multicoll. 
#cor(sampledf)
#come back to this
  
```
## Full model diagnostics
``` {r full model}
#full model LL 
logLik(logit_model)

#Diagnostic plots
plot(logit_model)
```
## Variable selection

```{r vars aic}
#AIC vs. BIC var selection criteria
#Akaike Info Criterion
final_aic<-bind_rows(dead, alive) |>
  dplyr::select(-Survival.Months)

stepAIC(logit_model, trace=0, k = 2)
#model determined by AIC criterion: minimize AIC for better model fit
logit_aic <- glm(Status ~ Age + Race + Marital.Status + X6th.Stage + 
    differentiate + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + 
    Reginol.Node.Positive, family = binomial(), data = final_aic)
#log likelihood
logLik(logit_aic)
vif(logit_aic)
#confusion matrix
predictions_aic <- predict(logit_aic, type = "response")
final_aic$predicted_classes <- ifelse(predictions_aic > 0.5, 1, 0)
confusionMatrix(as.factor(final_aic$predicted_classes), as.factor(final_aic$Status))
```
```{r bic}
#Bayesian Info Criterion
final_bic<-bind_rows(dead, alive) |>
  dplyr::select(-Survival.Months)

stepAIC(logit_model, trace=0, k = log(nrow(sampledf)))
#model determined by BIC criterion: minimize BIC for model fit
logit_bic <- glm(Status ~ differentiate + Progesterone.Status + 
    Regional.Node.Examined + Reginol.Node.Positive, family = binomial(), 
    data = final_bic)
#log likelihood
logLik(logit_bic)


#confusion matrix
predictions_bic <- predict(logit_bic, type = "response")
final_bic$predicted_classes <- ifelse(predictions_bic > 0.5, 1, 0)
confusionMatrix(as.factor(final_bic$predicted_classes), as.factor(final_bic$Status))
```

```{r diagnostics}
#make diagnostic plots to address outlying observations and assumptions
#Diagnostic plot for AIC model
plot(logit_aic)
```
```{r diagnostics 2}
#Diagnostic plot for BIC model
plot(logit_bic)


#refit models using sensitivity-specificity plots and then check performance and sensitivity/specificity
```

we can then pick and interpret the best final model
```{r}
```


## Use the balanced model for regression
```{r}
# Load required libraries
library(tidyverse)
library(survival)
library(caret)
library(car)  # For VIF checks
library(MASS) # For stepwise variable selection
library(glmnet)

# Load the balanced dataset
balance_data <- read.csv("project_2_numdata.csv")

# View dataset structure (optional)
str(balance_data)
view(balance_data)

# Summary statistics for numerical variables
summary(balance_data)

# Visualize distributions of numerical variables
numerical_vars <- balance_data[, sapply(balance_data, is.numeric)]

# Histograms
par(mfrow = c(2, 2))  # Arrange multiple plots
for (col in colnames(numerical_vars)) {
  hist(numerical_vars[[col]], main = paste("Histogram of", col), xlab = col, col = "lightblue")
}

# Density plots for more detailed distributions
for (col in colnames(numerical_vars)) {
  plot(density(numerical_vars[[col]], na.rm = TRUE), main = paste("Density Plot of", col), xlab = col)
}


```



```{r}
balance_data$race <- as.factor(balance_data$race)
balance_data$marital_status <- as.factor(balance_data$marital_status)
balance_data$a_stage <- as.factor(balance_data$a_stage)

# Frequency tables
categorical_vars <- balance_data[, sapply(balance_data, is.factor)]

for (col in colnames(categorical_vars)) {
  cat("Frequency of", col, ":\n")
  print(table(categorical_vars[[col]]))
}

# Bar plots
par(mfrow = c(2, 2))
for (col in colnames(categorical_vars)) {
  barplot(table(categorical_vars[[col]]), main = paste("Barplot of", col), col = "lightgreen")
}

```

```{r}

library(GGally)

# Correlation matrix
cor_matrix <- cor(balance_data[, sapply(balance_data, is.numeric)], use = "complete.obs")
print(cor_matrix)

# Visualize correlations
library(corrplot)
corrplot(cor_matrix, method = "circle", main = "Correlation Matrix")

# Pairwise scatterplots for key numerical variables
pairs(numerical_vars, main = "Pairwise Scatterplots")

ggpairs(numerical_vars, 
        title = "Pairwise Scatterplots for Numerical Variables",
        upper = list(continuous = wrap("cor", size = 4)), # Add correlation in upper triangle
        lower = list(continuous = "smooth"),             # Scatterplot with smooth line
        diag = list(continuous = "densityDiag"))         # Density plot on diagonal
```

```{r}
# Boxplot to examine relationship between a numerical variable and Status
boxplot(balance_data$tumor_size ~ balance_data$status, main = "Tumor Size by Status", xlab = "Status", ylab = "Tumor Size")

# Barplot for categorical variables vs. Status
for (col in colnames(categorical_vars)) {
  table_by_status <- table(balance_data[[col]], balance_data$status)
  barplot(table_by_status, beside = TRUE, main = paste(col, "by Status"), col = c("lightblue", "pink"))
}
```



```{r}
# Logistic Regression Full Model
logit_model_balance <- glm(status ~ ., data = balance_data, family = binomial())

# Model Summary
summary(logit_model_balance)

# Predictions
predictions <- predict(logit_model_balance, type = "response")

# Classify Predictions
balance_data$predicted_classes <- ifelse(predictions > 0.5, 1, 0)

# Ensure consistent factor levels
balance_data$predicted_classes <- factor(balance_data$predicted_classes, levels = c(0, 1))
balance_data$status <- factor(balance_data$status, levels = c(0, 1))

# Confusion Matrix
confusionMatrix(balance_data$predicted_classes, balance_data$status)

# Variance Inflation Factor (VIF) to detect multicollinearity
library(car)
#vif(logit_model_balance)

# Correlation Matrix
cor_matrix <- cor(balance_data[, sapply(balance_data, is.numeric)], use = "complete.obs")
print(cor_matrix)

# Log-Likelihood of the Full Model
logLik(logit_model_balance)

# Diagnostic Plots
par(mfrow = c(2, 2))
plot(logit_model_balance)
par(mfrow = c(1, 1))  # Reset graphical parameters
```



## Interaction models!!!

```{r}
library(fastDummies)

# Create dummy variables for categorical variables
balance_data_dummies <- dummy_cols(
  balance_data,
  select_columns = c("race", "marital_status", "x6th_stage", "a_stage"),
  remove_first_dummy = TRUE,  # Avoid multicollinearity by removing the first dummy
  remove_selected_columns = TRUE  # Remove original columns
)

# View the updated dataset
head(balance_data_dummies)
```



```{r} 
library(glmnet)

# Load required libraries
library(MASS)

# Generate all two-way interactions
formula_with_interactions <- status ~ 
  (age + grade + tumor_size + estrogen_status + progesterone_status + 
   regional_node_examined + reginol_node_positive)^2+  + 
   marital_status_Single + marital_status_Widowed + marital_status_Separated + 
   x6th_stage_IIB + x6th_stage_IIIA + x6th_stage_IIIB + x6th_stage_IIIC + 
   a_stage_regional
# Fit the logistic regression model with the updated formula
interaction_model <- glm(formula_with_interactions, data = balance_data_dummies, family = binomial())

# View the summary of the model
summary(interaction_model)
```


```{r}
# Perform Stepwise Selection (AIC)
aic_forward_model <- stepAIC(interaction_model, direction = "forward", trace = FALSE)
aic_backward_model <- stepAIC(interaction_model, direction = "backward", trace = FALSE)
aic_both_model <- stepAIC(interaction_model, direction = "both", trace = FALSE)

# Perform Stepwise Selection (BIC)
bic_forward_model <- stepAIC(interaction_model, direction = "forward", trace = FALSE, k = log(nrow(balance_data)))
bic_backward_model <- stepAIC(interaction_model, direction = "backward", trace = FALSE, k = log(nrow(balance_data)))
bic_both_model <- stepAIC(interaction_model, direction = "both", trace = FALSE, k = log(nrow(balance_data)))


```

```{r}
# Extract Coefficients for All Models
extract_coefficients <- function(model, model_name) {
  summary <- summary(model)
  coef <- as.data.frame(summary$coefficients)
  coef$Term <- rownames(coef)
  coef$Model <- model_name
  return(coef)
}

# Create Coefficient Tables
aic_forward_coef <- extract_coefficients(aic_forward_model, "AIC Forward")
aic_backward_coef <- extract_coefficients(aic_backward_model, "AIC Backward")
aic_both_coef <- extract_coefficients(aic_both_model, "AIC Both")
bic_forward_coef <- extract_coefficients(bic_forward_model, "BIC Forward")
bic_backward_coef <- extract_coefficients(bic_backward_model, "BIC Backward")
bic_both_coef <- extract_coefficients(bic_both_model, "BIC Both")

# Merge Results for AIC Models
aic_combined <- merge(
  aic_forward_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  aic_backward_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  by = "Term", all = TRUE, suffixes = c("_AIC_Forward", "_AIC_Backward")
)
aic_combined <- merge(
  aic_combined,
  aic_both_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  by = "Term", all = TRUE
)
colnames(aic_combined)[5:8] <- paste0(colnames(aic_combined)[5:8], "_AIC_Both")

# Merge Results for BIC Models
bic_combined <- merge(
  bic_forward_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  bic_backward_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  by = "Term", all = TRUE, suffixes = c("_BIC_Forward", "_BIC_Backward")
)
bic_combined <- merge(
  bic_combined,
  bic_both_coef[, c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")],
  by = "Term", all = TRUE
)
colnames(bic_combined)[5:8] <- paste0(colnames(bic_combined)[5:8], "_BIC_Both")

# Combine AIC and BIC Results
comparison_table <- merge(aic_combined, bic_combined, by = "Term", all = TRUE)

# Replace NA with 0 for consistency (optional)
comparison_table[is.na(comparison_table)] <- 0
library(writexl)   
# Export the Comparison Table to Excel
write_xlsx(comparison_table, "aic_bic_stepwise_comparison.xlsx")

# Confirm Export
cat("The stepwise comparison table has been successfully written to 'aic_bic_stepwise_comparison.xlsx'")

```

```{r}
library(modelsummary)
models <- list(
  "AIC Forward" = aic_forward_model,
  "AIC Backward" = aic_backward_model,
  "AIC Stepwise" = aic_both_model,
  "BIC Forward" = bic_forward_model,
  "BIC Backward" = bic_backward_model,
  "BIC Stepwise" = bic_both_model
)
modelsummary(models, output = "markdown")

```


```{r}
library(pROC) 
# Function for Model Evaluation
evaluate_model <- function(model, data, model_name) {
  # Log-Likelihood
  log_likelihood <- logLik(model)
  cat("Log-Likelihood for", model_name, ":", log_likelihood, "\n")
  
  # Predictions
  predictions <- predict(model, type = "response")
  data$predicted_classes <- ifelse(predictions > 0.5, 1, 0)
  
  # Ensure Consistent Factor Levels
  data$predicted_classes <- factor(data$predicted_classes, levels = c(0, 1))
  data$status <- factor(data$status, levels = c(0, 1))
  
  # Confusion Matrix
  conf_matrix <- confusionMatrix(data$predicted_classes, data$status)
  cat("Confusion Matrix for", model_name, ":\n")
  print(conf_matrix)
  
  # AUC-ROC
  roc_curve <- roc(as.numeric(as.character(data$status)), predictions)
  auc_value <- auc(roc_curve)
  cat("AUC for", model_name, ":", auc_value, "\n")
  
  return(list(conf_matrix = conf_matrix, auc = auc_value, roc_curve = roc_curve))
}


# Evaluate Models
cat("\n### Evaluating AIC Models ###\n")
aic_forward_results <- evaluate_model(aic_forward_model, balance_data, "AIC Forward")
aic_backward_results <- evaluate_model(aic_backward_model, balance_data, "AIC Backward")
aic_both_results <- evaluate_model(aic_both_model, balance_data, "AIC Stepwise")

cat("\n### Evaluating BIC Models ###\n")
bic_forward_results <- evaluate_model(bic_forward_model, balance_data, "BIC Forward")
bic_backward_results <- evaluate_model(bic_backward_model, balance_data, "BIC Backward")
bic_both_results <- evaluate_model(bic_both_model, balance_data, "BICS tepwise")

# Plot ROC Curves for All Models
plot(aic_forward_results$roc_curve, col = "blue", lwd = 2, main = "ROC Curves for All Models")
lines(aic_backward_results$roc_curve, col = "green", lwd = 2)
lines(aic_both_results$roc_curve, col = "cyan", lwd = 2)
lines(bic_forward_results$roc_curve, col = "red", lwd = 2)
lines(bic_backward_results$roc_curve, col = "orange", lwd = 2)
lines(bic_both_results$roc_curve, col = "purple", lwd = 2)

legend("bottomright", 
       legend = c("AIC Forward", "AIC Backward", "AIC Both", 
                  "BIC Forward", "BIC Backward", "BIC Both"),
       col = c("blue", "green", "cyan", "red", "orange", "purple"), 
       lwd = 2)

```
```{r}
summary(bic_forward_model)
```

```{r}
summary(bic_backward_model)
plot(bic_backward_model)
```

Evaluate the performance of your modells). ls your model achieving similar performancebetween the majority race group “White" and the minority “Black" (or “Black” + “Other")? lfnot, could you try to improve the fairness (i.e., reducing the gap of prediction performancebetween the majority and minority) of your model(s)?
```{r}
library(caret)
library(pROC)

# Assuming 'race' is your race variable and 'model' is your trained model
evaluate_performance <- function(data, model, group_name) {
  # Subset data
  subset_data <- data[data$race == group_name, ]
  
  # Predictions
  predictions <- predict(bic_backward_model, newdata = subset_data, type = "response")
  predicted_classes <- ifelse(predictions > 0.5, 1, 0)
  
  # Confusion Matrix
  conf_matrix <- confusionMatrix(factor(predicted_classes), factor(subset_data$status))
  print(paste("Confusion Matrix for", group_name))
  print(conf_matrix)
  
  # AUC-ROC
  roc_curve <- roc(subset_data$status, predictions)
  auc_value <- auc(roc_curve)
  print(paste("AUC for", group_name, ":", auc_value))
}

# Evaluate for 'White' group
evaluate_performance(balance_data, bic_backward_model, "White")

# Evaluate for 'Black' group
evaluate_performance(balance_data, bic_backward_model, "Black")

# Evaluate for 'Black' + 'Other' group
evaluate_performance(balance_data, bic_backward_model, c("Black", "Other"))
```



# Survival Analysis
```{r Multicollinearity for survival}
cox_model <- coxph(Surv(survival_months, status) ~.,
                   data = balance_df)
vifs <- vif(cox_model)  
print(vifs)

balance_df <-
  balance_df|>
  dplyr::select(-t_stage,-n_stage,-reginol_node_positive)
```
remove t_stage，n_stage and reginol_node_positive due to higher VIF score

```{r variable dplyr::selection for survival}
variables <- names(balance_df)[!names(balance_df) %in% c("survival_months", "status")]
for (var in variables) {
  formula <- as.formula(paste("Surv(survival_months, status) ~", var))
  model <- coxph(formula, data = balance_df)
  print(var)
  print(summary(model))
}

balance_df<-
  balance_df|>
  mutate(Married = ifelse(marital_status == "Married", 1, 0)) |>
  dplyr::select(-marital_status)
```

```{r ph assumption test}
ph_model<- coxph(Surv(survival_months, status) ~.,
                   data = balance_df)
ph_test <- cox.zph(ph_model)
print(ph_test)

final<- balance_df|>
  dplyr::select(-progesterone_status, -estrogen_status)
```

```{r fullmodel}
fullmodel <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + log2(tumor_size+1) + log2(regional_node_examined+1) + a_stage_regional + Married, data = final)
summary(Surv_fullmodel)
```


```{r AIC auto }
nullmodel <- coxph(Surv(survival_months, status) ~ 1, data = final)

forward_model <- step(nullmodel, scope = list(lower = nullmodel, upper = fullmodel), direction = "forward", trace = FALSE)
backward_model <- step(fullmodel, direction = "backward", trace = FALSE) 
stepwise_model <- step(nullmodel, scope=list(lower=nullmodel, upper=fullmodel), direction="both", trace = FALSE)
forward_model
backward_model
stepwise_model
```
```{r BIC}
stepAIC(fullmodel, trace = 0, k = log(nrow(final)))

Bic_model <- coxph(Surv(survival_months, status) ~ age + x6th_stage + grade + log2(regional_node_examined+1), data = final)
Bic_model
```
```{r Interact model }
Inter_model <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + log2(tumor_size+1) + log2(regional_node_examined+1) + a_stage_regional+ Married + Married*log2(regional_node_examined+1), data = final)
summary(Inter_model)
Inter_model
```
```{r Parsimony model }
Par_model <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + 
                            log2(regional_node_examined+1),
                            data = final)
summary(Par_model)
Par_model
```

```{r lasso model for survival}
predictors <- model.matrix(~ age + race + x6th_stage + log(tumor_size +1) + log2(regional_node_examined+1) + a_stage_regional + Married - 1, data = phob)
predictors_scaled <- scale(predictors)
response <- with(phob, Surv(survival_months, as.numeric(status)))
cv_lasso <- cv.glmnet(predictors_scaled, response, family = "cox", alpha = 1, type.measure = "deviance", nfolds = 10)
plot(cv_lasso)
best_lambda <- cv_lasso$lambda.min
final_lasso_model <- glmnet(predictors_scaled, response, family = "cox", alpha = 1, lambda = best_lambda)
lasso_coefficients <- coef(final_lasso_model)
print(lasso_coefficients)
lasso_coefficients
```


```{r performance}
fullmodel$concordance
Bic_model$concordance
Inter_model$concordance
Par_model$concordance

AIC(fullmodel)
AIC(Bic_model)
AIC(Inter_model)
AIC(Par_model)

BIC(fullmodel)
BIC(Bic_model)
BIC(Inter_model)
BIC(Par_model)
```

```{r outliers }
dev_residuals <- residuals(Inter_model, type = "deviance")
plot(dev_residuals, main = "Deviance Residuals", ylab = "Residuals", xlab = "Index")
abline(h = c(-2, 2), col = "red", lty = 2) 

surv_fit <- survfit(Surv(survival_months, status) ~ 1, data = final)

plot(surv_fit, xlab = "Time (months)", ylab = "Survival Probability", 
     main = "Survival Curve for the Final Model", col = "blue", lwd = 2)

```

```{r risk prediction }
predicted_hazards <- predict(Inter_model, type = "risk")
final$predicted_scores = predicted_hazards
```

```{r auc and disparity check}

time_points <-  c(2,43,62,82,107)

compute_auc <- function(data, time_point) {
    roc_data <- survivalROC(Stime = data$survival_months, status = data$status,
                            marker = data$predicted_scores, predict.time = time_point, method = "KM")
    return(roc_data$AUC)
}

data_white <- final[final$race == "White", ]
data_black <- final[final$race == "Black", ]
data_black_other <- final[final$race %in% c("Black", "Other"), ]

auc_white <- sapply(time_points, function(t) compute_auc(data_white, t))
auc_black <- sapply(time_points, function(t) compute_auc(data_black, t))
auc_black_other <- sapply(time_points, function(t) compute_auc(data_black_other, t))

print("Time-dependent AUC for White group:")
print(auc_white)

print("Time-dependent AUC for black group:")
print(auc_black)

print("Time-dependent AUC for Black/Other group:")
print(auc_black_other)

```


```{r quentile disparity }
quantiles_white <- quantile(final$survival_months[final$race == "White"], probs = c(0, 0.25, 0.5, 0.75,1))
quantiles_black <- quantile(final$survival_months[final$race == "Black"], probs = c(0, 0.25, 0.5, 0.75,1))
quantiles_blackother <- quantile(final$survival_months[final$race %in% c("Black", "Other")], probs = c(0, 0.25, 0.5, 0.75,1))



data_white <- final[final$race == "White", ]
data_black <- final[final$race == "Black", ]
data_black_other <- final[final$race %in% c("Black", "Other"), ]

compute_auc_quantiles <- function(data, quantiles) {
    sapply(quantiles, function(t) {
        roc_data <- survivalROC(Stime = data$survival_months, status = data$status,
                                marker = data$predicted_scores, predict.time = t, method = "KM")
        return(roc_data$AUC)
    })
}

compute_auc_quantiles(data_white, quantiles_white)
compute_auc_quantiles(data_black, quantiles_black)
compute_auc_quantiles(data_black_other, quantiles_blackother)

```



