---
title: "Final"
date: "`r Sys.Date()`"
output: html_document
---

# Setup and Libraries
```{r library}
library(tidyverse)
library(janitor)
library(ggplot2)
library(GGally)
library(dplyr)
library(gridExtra)
library(survival)
library(car)
library(PerformanceAnalytics)
library(glmnet)
library(caret)
library(pROC)
library(MASS)
library(writexl)
library(corrplot)
library(survivalROC)
set.seed(1)
```

```{r}
# Load the dataset
project_2_data <- read_csv("Project_2_data.csv", na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    status = ifelse(status == "Dead", 1, 0),
    t_stage = case_when(t_stage == "T1" ~ 1, t_stage == "T2" ~ 2, t_stage == "T3" ~ 3, t_stage == "T4" ~ 4),
    n_stage = case_when(n_stage == "N1" ~ 1, n_stage == "N2" ~ 2, n_stage == "N3" ~ 3),
    differentiate = case_when(
      differentiate == "Well differentiated" ~ 1,
      differentiate == "Moderately differentiated" ~ 2,
      differentiate == "Poorly differentiated" ~ 3,
      differentiate == "Undifferentiated" ~ 4
    ),
    grade = as.numeric(as.character(grade)),
    a_stage_regional = ifelse(a_stage == "Regional", 1, 0),
    estrogen_status = ifelse(estrogen_status == "Positive", 1, 0),
    progesterone_status = ifelse(progesterone_status == "Positive", 1, 0)
  ) |>
  stats::na.omit()

# Check missingness and descriptive statistics
summary(project_2_data)

```

```{r}
# Log transformations
project_2_data <- project_2_data |>
  mutate(
    log_tumor_size = log(tumor_size + 1),
    log_regional_node_examined = log(regional_node_examined + 1)
  )

# Visualize transformations
ggplot(project_2_data, aes(x = sqrt(tumor_size))) +
  geom_histogram(fill = "green", color = "black") +
  labs(title = "Log-Transformed Regional Nodes Examined")

ggplot(project_2_data, aes(x = log_tumor_size)) +
  geom_histogram(fill = "blue", color = "black") +
  labs(title = "Log-Transformed Tumor Size")     #Best

ggplot(project_2_data, aes(x = sqrt(regional_node_examined))) +
  geom_histogram(fill = "green", color = "black") +
  labs(title = "Log-Transformed Regional Nodes Examined")

ggplot(project_2_data, aes(x = log_regional_node_examined)) +
  geom_histogram(fill = "green", color = "black") +
  labs(title = "Log-Transformed Regional Nodes Examined")    #Best

```

```{r}
# Correlation matrix and multicollinearity
cor_matrix <- cor(project_2_data[, sapply(project_2_data, is.numeric)], use = "complete.obs")
corrplot(cor_matrix, method = "circle")

```

```{r}
# Balance the dataset
dead <- project_2_data |> filter(status == 1)
alive <- project_2_data |> filter(status == 0) |> sample_n(nrow(dead))
balance_data <- bind_rows(dead, alive)

# Visualize balance
ggplot(balance_data, aes(x = status)) + geom_bar() + labs(title = "Balanced Dataset")

```
logistic without interaction
```{r}
# Full logistic regression model
logit_model <- glm(status ~ ., data = balance_data, family = binomial())
summary(logit_model)

# Variable selection: AIC and BIC
aic_model <- stepAIC(logit_model, trace = FALSE)
bic_model <- stepAIC(logit_model, k = log(nrow(balance_data)), trace = FALSE)

# Model performance
evaluate_model <- function(model, data, model_name) {
  predictions <- predict(model, type = "response")
  data$predicted_classes <- ifelse(predictions > 0.5, 1, 0)
  conf_matrix <- confusionMatrix(factor(data$predicted_classes), factor(data$status))
  roc_curve <- roc(data$status, predictions)
  list(
    LogLikelihood = logLik(model),
    ConfusionMatrix = conf_matrix,
    AUC = auc(roc_curve)
  )
}

logit_results_aic <- evaluate_model(aic_model, balance_data, "AIC Model")
logit_results_bic <- evaluate_model(bic_model, balance_data, "BIC Model")

```
interation logistic!!!!
```{r}

library(fastDummies)

# Create dummy variables for categorical variables
balance_data_dummies <- dummy_cols(
  balance_data,
  select_columns = c("race", "marital_status", "x6th_stage", "a_stage"),
  remove_first_dummy = TRUE,  # Avoid multicollinearity by removing the first dummy
  remove_selected_columns = TRUE  # Remove original columns
)

# View the updated dataset
head(balance_data_dummies)

# Generate all two-way interactions
formula_with_interactions <- status ~ 
  (age + grade + tumor_size + estrogen_status + progesterone_status + 
   regional_node_examined + reginol_node_positive)^2 + 
   marital_status_Single + marital_status_Widowed + marital_status_Separated + 
   x6th_stage_IIB + x6th_stage_IIIA + x6th_stage_IIIB + x6th_stage_IIIC + 
   a_stage_regional

# Fit the logistic regression model with interaction terms
interaction_model <- glm(formula_with_interactions, data = balance_data_dummies, family = binomial())

# Summary of the model
summary(interaction_model)

```
Model Selection with Stepwise Approaches
```{r}
# Perform Stepwise Selection (AIC)
aic_forward_model <- stepAIC(interaction_model, direction = "forward", trace = FALSE)
aic_backward_model <- stepAIC(interaction_model, direction = "backward", trace = FALSE)
aic_both_model <- stepAIC(interaction_model, direction = "both", trace = FALSE)

# Perform Stepwise Selection (BIC)
bic_forward_model <- stepAIC(interaction_model, direction = "forward", trace = FALSE, k = log(nrow(balance_data)))
bic_backward_model <- stepAIC(interaction_model, direction = "backward", trace = FALSE, k = log(nrow(balance_data)))
bic_both_model <- stepAIC(interaction_model, direction = "both", trace = FALSE, k = log(nrow(balance_data)))

```
Coefficient Extraction and Export
```{r}
# Extract coefficients from models
extract_coefficients <- function(model, model_name) {
  coef <- as.data.frame(summary(model)$coefficients)
  coef$Term <- rownames(coef)
  coef$Model <- model_name
  return(coef)
}

# Combine coefficients from AIC and BIC models
aic_combined <- merge(
  extract_coefficients(aic_forward_model, "AIC Forward"),
  extract_coefficients(aic_backward_model, "AIC Backward"),
  by = "Term", all = TRUE, suffixes = c("_AIC_Forward", "_AIC_Backward")
)

aic_combined <- merge(
  aic_combined,
  extract_coefficients(aic_both_model, "AIC Both"),
  by = "Term", all = TRUE
)

bic_combined <- merge(
  extract_coefficients(bic_forward_model, "BIC Forward"),
  extract_coefficients(bic_backward_model, "BIC Backward"),
  by = "Term", all = TRUE, suffixes = c("_BIC_Forward", "_BIC_Backward")
)

bic_combined <- merge(
  bic_combined,
  extract_coefficients(bic_both_model, "BIC Both"),
  by = "Term", all = TRUE
)

# Combine AIC and BIC results into one table
comparison_table <- merge(aic_combined, bic_combined, by = "Term", all = TRUE)

# Replace NA with 0 for consistency
comparison_table[is.na(comparison_table)] <- 0

# Export the table
write_xlsx(comparison_table, "aic_bic_stepwise_comparison.xlsx")
cat("Comparison table exported to 'aic_bic_stepwise_comparison.xlsx'")

```
Model Evaluation
```{r}
evaluate_model <- function(model, data, model_name) {
  # Log-Likelihood
  log_likelihood <- logLik(model)
  cat("Log-Likelihood for", model_name, ":", log_likelihood, "\n")
  
  # Predictions
  predictions <- predict(model, type = "response")
  
  # Check for valid predictions
  if (any(!is.finite(predictions))) {
    cat("Invalid predictions detected for", model_name, "\n")
    return(NULL)
  }
  
  # Classify predictions
  data$predicted_classes <- ifelse(predictions > 0.5, 1, 0)
  
  # Ensure Consistent Factor Levels
  data$predicted_classes <- factor(data$predicted_classes, levels = c(0, 1))
  data$status <- factor(data$status, levels = c(0, 1))
  
  # Confusion Matrix
  conf_matrix <- confusionMatrix(data$predicted_classes, data$status)
  cat("Confusion Matrix for", model_name, ":\n")
  print(conf_matrix)
  
  # AUC-ROC Calculation
  if (any(is.na(data$status)) || any(is.na(predictions))) {
    cat("Invalid data for ROC calculation in", model_name, "\n")
    return(NULL)
  }
  
  roc_curve <- tryCatch({
    roc(as.numeric(as.character(data$status)), predictions)
  }, error = function(e) {
    cat("Error calculating ROC for", model_name, ":", e$message, "\n")
    return(NULL)
  })
  
  if (is.null(roc_curve)) {
    cat("ROC calculation failed for", model_name, "\n")
    return(NULL)
  }
  
  auc_value <- auc(roc_curve)
  cat("AUC for", model_name, ":", auc_value, "\n")
  
  return(list(conf_matrix = conf_matrix, auc = auc_value, roc_curve = roc_curve))
}

# Evaluate AIC and BIC Models
aic_forward_results <- evaluate_model(aic_forward_model, balance_data, "AIC Forward")
aic_backward_results <- evaluate_model(aic_backward_model, balance_data, "AIC Backward")
aic_both_results <- evaluate_model(aic_both_model, balance_data, "AIC Stepwise")
bic_forward_results <- evaluate_model(bic_forward_model, balance_data, "BIC Forward")
bic_backward_results <- evaluate_model(bic_backward_model, balance_data, "BIC Backward")
bic_both_results <- evaluate_model(bic_both_model, balance_data, "BIC Stepwise")

# Plot ROC Curves

plot(aic_forward_results$roc_curve, col = "blue", lwd = 2, main = "ROC Curves for All Models")
lines(aic_backward_results$roc_curve, col = "green", lwd = 2)
lines(aic_both_results$roc_curve, col = "cyan", lwd = 2)
lines(bic_forward_results$roc_curve, col = "red", lwd = 2)
lines(bic_backward_results$roc_curve, col = "orange", lwd = 2)
lines(bic_both_results$roc_curve, col = "purple", lwd = 2)

legend("bottomright", 
       legend = c("AIC Forward", "AIC Backward", "AIC Both", "BIC Forward", "BIC Backward", "BIC Both"),
       col = c("blue", "green", "cyan", "red", "orange", "purple"), 
       lwd = 2)

```
Fairness Evaluation
```{r}
evaluate_performance <- function(data, model, group_name) {
  subset_data <- data |> filter(race %in% group_name)
  predictions <- predict(model, newdata = subset_data, type = "response")
  predicted_classes <- ifelse(predictions > 0.5, 1, 0)
  conf_matrix <- confusionMatrix(factor(predicted_classes), factor(subset_data$status))
  roc_curve <- roc(subset_data$status, predictions)
  
  list(
    ConfusionMatrix = conf_matrix,
    AUC = auc(roc_curve)
  )
}

# Evaluate performance for different racial groups
white_results <- evaluate_performance(balance_data, bic_backward_model, "White")
black_results <- evaluate_performance(balance_data, bic_backward_model, "Black")
black_other_results <- evaluate_performance(balance_data, bic_backward_model, c("Black", "Other"))

list(
  White = white_results,
  Black = black_results,
  Black_Other = black_other_results
)

```


Survival Analysis Models
```{r}

### Multicollinearity Analysis

cox_model <- coxph(Surv(survival_months, status) ~ .-a_stage -differentiate, data = balance_data)
vifs <- vif(cox_model)
print(vifs)

# Remove variables with high VIF scores
balance_data <- balance_data |> 
  dplyr::select(-t_stage, -n_stage, -reginol_node_positive)

```
Variable Selection
```{r}
# Evaluate each variable individually
variables <- names(balance_data)[!names(balance_data) %in% c("survival_months", "status")]
for (var in variables) {
  formula <- as.formula(paste("Surv(survival_months, status) ~", var))
  model <- coxph(formula, data = balance_data)
  print(var)
  print(summary(model))
}

# Recode and select relevant variables
balance_data <- balance_data |> 
  mutate(Married = ifelse(marital_status == "Married", 1, 0)) |> 
  dplyr::select(-marital_status)

```
Assumption Testing
```{r}
ph_model <- coxph(Surv(survival_months, status) ~ ., data = balance_data)
ph_test <- cox.zph(ph_model)
print(ph_test)

# Remove variables that violate assumptions
final <- balance_data |> 
  dplyr::select(-progesterone_status, -estrogen_status)

```
Model Fitting
```{r}
# Full Cox Model
fullmodel <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + 
                   log2(tumor_size + 1) + log2(regional_node_examined + 1) + 
                   a_stage_regional + Married, data = final)
summary(fullmodel)


```
Model Selection (AIC & BIC)
```{r}
# Stepwise Model Selection (AIC)
nullmodel <- coxph(Surv(survival_months, status) ~ 1, data = final)
forward_model <- step(nullmodel, scope = list(lower = nullmodel, upper = fullmodel), direction = "forward", trace = FALSE)
backward_model <- step(fullmodel, direction = "backward", trace = FALSE)
stepwise_model <- step(nullmodel, scope = list(lower = nullmodel, upper = fullmodel), direction = "both", trace = FALSE)
```

Interaction and Parsimony Models
```{r}
# Interaction Model
Inter_model <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + 
                     log2(tumor_size + 1) + log2(regional_node_examined + 1) + 
                     a_stage_regional + Married + 
                     Married * log2(regional_node_examined + 1), data = final)
summary(Inter_model)

# Parsimony Model
Par_model <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + 
                   log2(regional_node_examined + 1), data = final)
summary(Par_model)

```
Performance Evaluation
```{r}
# Compare Models
fullmodel$concordance
Inter_model$concordance
Par_model$concordance

# Compare AIC and BIC
AIC(fullmodel)
AIC(Inter_model)
AIC(Par_model)

BIC(fullmodel)
BIC(Inter_model)
BIC(Par_model)
```
```{r}
dev_residuals <- residuals(Inter_model, type = "deviance")
plot(dev_residuals, main = "Deviance Residuals", ylab = "Residuals", xlab = "Index")
abline(h = c(-3, 3), col = "red", lty = 3)
abline(h = c(-2, 2), col = "blue", lty = 3) 

surv_fit <- survfit(Surv(survival_months, status) ~ 1, data = final)

plot(surv_fit, xlab = "Time (months)", ylab = "Survival Probability", 
     main = "Survival Curve for the Final Model", col = "blue", lwd = 2)

grid()
```


Comparison and Fairness Check
```{r}
predicted_hazards <- predict(Inter_model, type = "risk")
final$predicted_scores = predicted_hazards

# Compute survival quantiles based on overall quantiles
time_points <-  c(2,43,62,82,107)

compute_auc <- function(data, time_point) {
    roc_data <- survivalROC(Stime = data$survival_months, status = data$status,
                            marker = data$predicted_scores, predict.time = time_point, method = "KM")
    return(roc_data$AUC)
}

data_white <- final[final$race == "White", ]
data_black <- final[final$race == "Black", ]
data_black_other <- final[final$race %in% c("Black", "Other"), ]

auc_white <- sapply(time_points, function(t) compute_auc(data_white, t))
auc_black <- sapply(time_points, function(t) compute_auc(data_black, t))
auc_black_other <- sapply(time_points, function(t) compute_auc(data_black_other, t))

print("Time-dependent AUC for White group:")
print(auc_white)

print("Time-dependent AUC for black group:")
print(auc_black)

print("Time-dependent AUC for Black/Other group:")
print(auc_black_other)


# Compute survival quantiles for each race group based on race own quantiles
quantiles_white <- quantile(final$survival_months[final$race == "White"], probs = c(0, 0.25, 0.5, 0.75, 1))
quantiles_black <- quantile(final$survival_months[final$race == "Black"], probs = c(0, 0.25, 0.5, 0.75, 1))
quantiles_blackother <- quantile(final$survival_months[final$race %in% c("Black", "Other")], probs = c(0, 0.25, 0.5, 0.75, 1))

# Subset data by race groups
data_white <- final[final$race == "White", ]
data_black <- final[final$race == "Black", ]
data_black_other <- final[final$race %in% c("Black", "Other"), ]

# Function to compute AUC at specific quantiles
compute_auc_quantiles <- function(data, quantiles) {
    sapply(quantiles, function(t) {
        roc_data <- survivalROC(Stime = data$survival_months, status = data$status,
                                marker = data$predicted_scores, predict.time = t, method = "KM")
        return(roc_data$AUC)
    })
}

# Calculate AUC for each group at quantiles
auc_white <- compute_auc_quantiles(data_white, quantiles_white)
auc_black <- compute_auc_quantiles(data_black, quantiles_black)
auc_black_other <- compute_auc_quantiles(data_black_other, quantiles_blackother)

# Print results
cat("Time-dependent AUC for White group:\n")
print(auc_white)

cat("Time-dependent AUC for Black group:\n")
print(auc_black)

cat("Time-dependent AUC for Black/Other group:\n")
print(auc_black_other)


```

