---
title: "survival numerical"
author: "Shiying Wu"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(car)
library(ggplot2)
library("PerformanceAnalytics")
```


```{r data}
set.seed(1)
project_2_data = 
  read_csv("./data/project_2_data.csv", na = c("NA", "", ".")) |>
  janitor::clean_names()  |>
  mutate(status = ifelse(status == "Dead", 1, 0))
      

dead <- project_2_data |>
  filter(status == 1)

alive <- project_2_data |>
  filter(status == 0) |>
  sample_n(nrow(dead))

project_2_numdata<-
  bind_rows(dead, alive) |>
  mutate(
    t_stage = case_when(
    t_stage == "T1" ~ 1,
    t_stage == "T2" ~ 2,
    t_stage == "T3" ~ 3,
    t_stage == "T4" ~ 4,
    TRUE ~ NA_real_),
    n_stage = case_when(
    n_stage == "N1" ~ 1,
    n_stage == "N2" ~ 2,
    n_stage == "N3" ~ 3,
    TRUE ~ NA_real_),
    differentiate = case_when(
    differentiate == "Well differentiated" ~ 1,
    differentiate == "Moderately differentiated" ~ 2,
    differentiate == "Poorly differentiated" ~ 3,
    differentiate == "Undifferentiated" ~ 4,
    TRUE ~ NA_real_),
    grade = case_when(
    grade == "anaplastic; Grade IV" ~ 4,
    grade == "3" ~ 3,
    grade == "2" ~ 2,
    grade == "1" ~ 1,
    TRUE ~ NA_real_),
    a_stage_regional = ifelse(a_stage == "Regional", 1, 0),
    estrogen_status = ifelse(estrogen_status == "Positive", 1, 0),
    progesterone_status = ifelse(progesterone_status == "Positive", 1, 0)
    ) |>
  select(-a_stage)

```


```{r variable selection}
variables <- names(project_2_numdata)[!names(project_2_numdata) %in% c("survival_months", "status")]
for (var in variables) {
  formula <- as.formula(paste("Surv(survival_months, status) ~", var))
  model <- coxph(formula, data = project_2_numdata)
  print(var)
  print(summary(model))
}
```
only marriage 

```{r Multicollinearity}
project_2_numdata <-
  project_2_numdata|>
  mutate(Married = ifelse(marital_status == "Married", 1, 0)) |>
  select(-differentiate, - marital_status)

cor_matrix <- cor(project_2_numdata[, sapply(project_2_numdata, is.numeric)])
print(cor_matrix)


cortable<-project_2_numdata|>
  select(-race,-x6th_stage)
cortable$regional_node_examined
chart.Correlation(cortable, histogram=TRUE, pch=19)
```
We have to choice 1 between 
t_stage - tumor size
n stage - regional_node_possitive

```{r Multicollinearity continous}
muldata = project_2_numdata
  
  
cox_model <- coxph(Surv(survival_months, status) ~.,
                   data = muldata)

vifs <- vif(cox_model)  
print(vifs)
```


```{r model}
final = muldata |>
  select(-t_stage, -n_stage)
  
cox_model <- coxph(Surv(survival_months, status) ~.,
                   data = final)
vifs <- vif(cox_model)
print(vifs)
```
```{r phtest}
library(survival)
ph_test <- cox.zph(cox_model)
print(ph_test)
plot(ph_test) 

```

The Cox model assumes that hazard ratios are constant over time. A non-significant p-value (p > 0.05) indicates that the PH assumption holds. As GLOBAL 48.732 15 1.9e-05, the model did not meet the assumption, as same as estrogen_status, progesterone_status, and a_stage_regional. We need further improve our model.

```{r phtest improved}
finalmodel <- coxph(Surv(survival_months, status) ~ age + race + x6th_stage + grade + 
                            tumor_size + regional_node_examined + a_stage_regional + Married,
                            data = final)


ph_test <- cox.zph(finalmodel)
print(ph_test)
plot(ph_test) 

cox_model$concordance
finalmodel$concordance
```

```{r outliers}
dev_residuals <- residuals(cox_model, type = "deviance")
plot(dev_residuals, main = "Deviance Residuals", ylab = "Residuals", xlab = "Index")
abline(h = c(-2, 2), col = "red", lty = 2) 

surv_fit <- survfit(Surv(survival_months, status) ~ 1, data = final)

plot(surv_fit, xlab = "Time (months)", ylab = "Survival Probability", 
     main = "Survival Curve for the Final Model", col = "blue", lwd = 2)

grid()

```
