---
title: "survival numerical"
author: "Shiying Wu"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(car)
library(ggplot2)
library("PerformanceAnalytics")
```


```{r data}
project_2_numdata = 
  read_csv("./data/project_2_data.csv", na = c("NA", "", ".")) |>
  janitor::clean_names()  |>
  mutate(status = ifelse(status == "Dead", 1, 0),
    t_stage = case_when(
    t_stage == "T1" ~ 1,
    t_stage == "T2" ~ 2,
    t_stage == "T3" ~ 3,
    t_stage == "T4" ~ 4,
    TRUE ~ NA_real_),
    n_stage = case_when(
    n_stage == "N1" ~ 1,
    n_stage == "N2" ~ 2,
    n_stage == "N3" ~ 3,
    TRUE ~ NA_real_),
    x6th_stage_num = case_when(
    x6th_stage == "IIA" ~ 1,
    x6th_stage == "IIB" ~ 2,
    x6th_stage == "IIIA" ~ 3,
    x6th_stage == "IIIB" ~ 4,
    x6th_stage == "IIIC" ~ 5,
    TRUE ~ NA_real_),
    differentiate = case_when(
    differentiate == "Well differentiated" ~ 1,
    differentiate == "Moderately differentiated" ~ 2,
    differentiate == "Poorly differentiated" ~ 3,
    differentiate == "Undifferentiated" ~ 4,
    TRUE ~ NA_real_),
    grade = case_when(
    grade == "anaplastic; Grade IV" ~ 4,
    grade == "3" ~ 3,
    grade == "2" ~ 2,
    grade == "1" ~ 1,
    TRUE ~ NA_real_),
    a_stage_regional = ifelse(a_stage == "Regional", 1, 0),
    estrogen_status = ifelse(estrogen_status == "Positive", 1, 0),
    progesterone_status = ifelse(progesterone_status == "Positive", 1, 0)
    ) |>
  select(-a_stage)

surv_object <- Surv(time = project_2_numdata$survival_months, event = project_2_numdata$status)
```


```{r variable selection}
variables <- names(project_2_numdata)[!names(project_2_numdata) %in% c("survival_months", "status")]
for (var in variables) {
  formula <- as.formula(paste("Surv(survival_months, status) ~", var))
  model <- coxph(formula, data = project_2_numdata)
  print(var)
  print(summary(model))
}
```
Based on the univariate Cox model results, the final model should include statistically significant and clinically relevant variables. These include age (p = 0.0007, HR = 1.016), race (p < 0.001, indicating disparities), marital status (significant for Married and Separated), t_stage (p < 2e-16, HR = 1.597), n_stage (p < 2e-16, HR = 2.151), differentiate (p < 2e-16, HR = 1.92), grade (p-value: < 2e-16, HR = 1.921), x6th_stage(p < 2e-16, HR = 1.578) tumor size (p < 2e-16, HR = 1.013), estrogen status (p < 2e-16, HR = 0.273), progesterone status (p < 2e-16, HR = 0.384), regional node positive (p < 2e-16, HR = 1.081), and a_stage (regional) (p < 2e-08, HR = 0.32). Variables like Single and Widowed marital status (p > 0.05) should be excluded. But we need further check multicollinearity and other for final dicision

```{r Multicollinearity}

cor_matrix <- cor(project_2_numdata[, sapply(project_2_numdata, is.numeric)])
print(cor_matrix)


cortable<-project_2_numdata|>
  select(-race, -marital_status,-x6th_stage)
cortable$regional_node_examined
chart.Correlation(cortable, histogram=TRUE, pch=19)
```
We have to choice 1 between 
differentiate - grade
t stage  - tumor size
n stage - regional_node_possitive - x6th_stage_num

```{r Multicollinearity continous}
muldata = project_2_numdata |>
  select(-differentiate, -x6th_stage_num)
  
cox_model <- coxph(Surv(survival_months, status) ~.,
                   data = muldata)

vifs <- vif(cox_model)  
print(vifs)
```


The VIF analysis reveals significant multicollinearity for n_stage (VIF = 10.94, GVIF^(1/(2Df)) =3.31), x6th_stage_num (VIF = 1.56, GVIF^(1/(2Df)) =1.25), reginol_node_positive (VIF = 4.13, GVIF^(1/(2Df)) =2.03), and t_stage (VIF = 3.83, GVIF^(1/(2Df)) =1.96), indicating redundancy. Tumor_size shows moderate multicollinearity (VIF = 2.62, GVIF^(1/(2Df)) =1.62), while other variables have acceptable VIFs near 1. Variables with high multicollinearity should be reconsidered for exclusion.
also shows we need to remove. So I decided to remove t_stage, n_stage, reginol_node_positive, 

```{r model}
final = muldata |>
  select(-t_stage, -n_stage, -reginol_node_positive)
  
cox_model <- coxph(Surv(survival_months, status) ~.,
                   data = final)
vifs <- vif(cox_model)
print(vifs)
```
```{r phtest}
library(survival)
ph_test <- cox.zph(cox_model)
print(ph_test)
plot(ph_test) 

```

The Cox model assumes that hazard ratios are constant over time. A non-significant p-value (p > 0.05) indicates that the PH assumption holds. As GLOBAL 50.520 14 5.0e-06, the model did not meet the assumption, as same as estrogen_status, progesterone_status, and a_stage_regional. We need further improve our model.
```{r phtest improved}
finalmodel <- coxph(Surv(survival_months, status) ~ age + race + marital_status +
                            grade + tumor_size + 
                            regional_node_examined + x6th_stage,
                            data = final)


ph_test <- cox.zph(finalmodel)
print(ph_test)
plot(ph_test) 

cox_model$concordance
finalmodel$concordance
```


No Clear Trends: If the solid line remains flat (close to zero), it indicates that the PH assumption is satisfied for that variable.
Upward/Downward Trends: A visible trend or deviation indicates that the proportional hazards assumption may be violated for the corresponding variable, suggesting time-dependent effects.

A C-index of 0.716 suggests that the model has good discriminatory power, meaning it can correctly rank the survival times for about 71.6% of the pairs.The standard error is 0.01075, indicating a narrow range of variability in the concordance estimate, suggesting robust performance. 
The reduction in the C-index (from 73.9% to 71.6%) indicates a trade-off between model complexity and performance.

```{r outliers}
dev_residuals <- residuals(cox_model, type = "deviance")
plot(dev_residuals, main = "Deviance Residuals", ylab = "Residuals", xlab = "Index")
abline(h = c(-2, 2), col = "red", lty = 2) 

surv_fit <- survfit(Surv(survival_months, status) ~ 1, data = final)

plot(surv_fit, xlab = "Time (months)", ylab = "Survival Probability", 
     main = "Survival Curve for the Final Model", col = "blue", lwd = 2)

grid()

```
